---
# IMPORTANT: Change settings here, but DO NOT change the spacing.
# Remove comments and add values where applicable.
# The descriptions below should be self-explanatory

title: "Beyond Money: What Truly Buys Happiness Across the Globe"
#subtitle: "This will appear as Right Header"

documentclass: "elsarticle"

# --------- Thesis title (Optional - set to FALSE by default).
# You can move the details below around as you please.
Thesis_FP: FALSE
# Entry1: "An unbelievable study with a title spanning multiple lines."
# Entry2: "\\textbf{Some Guy}" # textbf for bold
# Entry3: "A thesis submitted toward the degree of Doctor of Philosophy"
# Uni_Logo: Tex/Logo.png # Place a logo in the indicated location (from your root, e.g. defaults to ~/Tex/Logo.png) and uncomment this line. Leave uncommented for no image
# Logo_width: 0.3 # If using a logo - use this to set width (size) of image
# Entry4: "Under the supervision of: \\vfill Prof. Joe Smith and Dr. Frank Smith"
# Entry5: "Stellenbosch University"
# Entry6: April 2020
# Entry7:
# Entry8:

# --------- Front Page
# Comment: ----- Follow this pattern for up to 5 authors
AddTitle: TRUE # Use FALSE when submitting to peer reviewed platform. This will remove author names.
Author1: "Linda Dube"  # First Author - note the thanks message displayed as an italic footnote of first page.
Ref1: "Stellenbosch Wniversity, Western Cape" # First Author's Affiliation
Email1: "23084103\\@sun.ac.za" # First Author's Email address



CorrespAuthor_1: False  # If corresponding author is author 3, e.g., use CorrespAuthor_3: TRUE



# ----- Manage headers and footers:
#BottomLFooter: $Title$
#BottomCFooter:
#TopLHeader: \leftmark # Adds section name at topleft. Remove comment to add it.
BottomRFooter: "\\footnotesize Page \\thepage" # Add a '#' before this line to remove footer.
addtoprule: TRUE
addfootrule: TRUE               # Use if footers added. Add '#' to remove line.

# --------- page margins:
margin: 2.3 # Sides
bottom: 2 # bottom
top: 2.5 # Top
HardSet_layout: TRUE # Hard-set the spacing of words in your document. This will stop LaTeX squashing text to fit on pages, e.g.
# This is done by hard-setting the spacing dimensions. Set to FALSE if you want LaTeX to optimize this for your paper.

# --------- Line numbers
linenumbers: FALSE # Used when submitting to journal

# ---------- References settings:
# You can download cls format here: https://www.zotero.org/ - simply search for your institution. You can also edit and save cls formats here: https://editor.citationstyles.org/about/
# Hit download, store it in Tex/ folder, and change reference below - easy.
bibliography: Tex/ref.bib       # Do not edit: Keep this naming convention and location.
csl: Tex/harvard-stellenbosch-university.csl # referencing format used.
# By default, the bibliography only displays the cited references. If you want to change this, you can comment out one of the following:
#nocite: '@*' # Add all items in bibliography, whether cited or not
# nocite: |  # add specific references that aren't cited
#  @grinold2000
#  @Someoneelse2010

# ---------- General:
RemovePreprintSubmittedTo: TRUE  # Removes the 'preprint submitted to...' at bottom of titlepage
Journal: "Journal of Finance"   # Journal that the paper will be submitting to, if RemovePreprintSubmittedTo is set to TRUE.
toc: FALSE                       # Add a table of contents
numbersections: TRUE             # Should sections (and thus figures and tables) be numbered?
fontsize: 11pt                  # Set fontsize
linestretch: 1.2                # Set distance between lines.
link-citations: TRUE            # This creates dynamic links to the papers in reference list.

### Adding additional latex packages:
# header-includes:
#    - \usepackage{colortbl} # Add additional packages here.

output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 3.5 # Adjust default figure sizes. This can also be done in the chunks of the text.
    fig_height: 3.5
    latex_engine: xelatex
abstract: "This report investigates the key drivers of happiness across global regions, using the Gallup World Poll data. Rather than focusing solely on income, we explore whether other dimensions—particularly Healthy Life Expectancy—are stronger predictors of well-being. Happiness is measured via the Ladder Score, which captures life satisfaction. Visual and descriptive analysis reveals significant variation in life expectancy and social support across regions. The report provides new insights into what truly contributes to happiness beyond material wealth."

---
<!-- First: Set your default preferences for chunk options: -->

<!-- If you want a chunk's code to be printed, set echo = TRUE. message = FALSE stops R printing ugly package loading details in your final paper too. I also suggest setting warning = FALSE and checking for warnings in R, else you might find ugly warnings in your paper. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.pos="H", fig.pos = 'H')

if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)

list.files("C:/Users/Katli Gabs/Documents/Data Science/27674193/27674193/Question1/Question1/code/Question1_Plots.R", full.names = T, recursive = T) %>% as.list() %>% walk(~source(.))
```


<!-- ############################## -->
<!-- # Start Writing here: -->
<!-- ############################## -->

# Introduction \label{Introduction}



The question of what truly drives human happiness has long fascinated social scientists. While many studies have emphasized the role of income in shaping life satisfaction, emerging perspectives challenge the idea that money alone can guarantee well-being. This report draws on data from the Gallup World Poll, which measures happiness scores across countries and ranks them based on individuals’ self-reported life evaluations, also known as Ladder Scores. These scores are influenced by six key factors: economic production, social support, healthy life expectancy, freedom, absence of corruption, and generosity. Though these components do not directly determine the total happiness score, they help explain differences in happiness across countries by comparing each nation to Dystopia—a hypothetical benchmark representing the lowest global averages across all factors.

In this analysis, we shift the focus from income to healthy life expectancy, asking: Could good health, rather than economic wealth, be the foundation of happiness? While it is often assumed that higher incomes lead to higher life satisfaction, past research, has shown that beyond a certain point, increases in income do not translate into lasting improvements in happiness. This report therefore investigates whether health-related well-being is a more consistent and meaningful driver of happiness across regions than financial wealth.


# Data Description
```{r}
library(tidyverse)

#  file path 
folder_path <- "C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/mock exam/Tex_Ex/23084103/Question 1/Tex_Ex/23084103_HAPPINESS/data/Happy"

# List all CSV files in the folder
csv_files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)

# Check: Print to confirm files are found
print(csv_files)

# Combine all CSV files into one data frame
combined_happy_data <- csv_files %>%
  map_dfr(read_csv)

# View the result
View(combined_happy_data)
```
```{r}


library(skimr)

skim(combined_happy_data)
```





To better understand the characteristics of the happiness data, we present a descriptive statistics table summarizing the key variables. The average ladder score, a proxy for life satisfaction, is approximately 5.53, with a standard deviation of 1.07, indicating moderate variation across regions. The average healthy life expectancy stands at nearly 65 years, suggesting relatively good health outcomes globally, though some variation is evident. Variables such as social support, freedom to make life choices, and perceptions of corruption exhibit diverse levels across observations. Overall, the data is complete with no missing values for the main variables analyzed, making it suitable for robust further modeling.

<!-- The following is a code chunk. It must have its own unique name (after the r), or no name. After the comma follows commands for R which are self-explanatory. By default, the code and messages will not be printed in your pdf, just the output: -->

1a)


```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Calculate average life expectancy per region
region_summary <- combined_happy_data %>%
  group_by(`Regional indicator`) %>%
  summarise(
    `Ladder score` = mean(`Ladder score`, na.rm = TRUE),
    upperwhisker = mean(upperwhisker, na.rm = TRUE),
    lowerwhisker = mean(lowerwhisker, na.rm = TRUE),
    `Healthy life expectancy` = mean(`Healthy life expectancy`, na.rm = TRUE)
  ) %>%
  arrange(`Healthy life expectancy`)

# Plot
ggplot(region_summary, aes(x = `Ladder score`, y = reorder(`Regional indicator`, `Healthy life expectancy`))) +
  geom_point(color = "blue", size = 2) +
  geom_errorbarh(aes(xmin = lowerwhisker, xmax = upperwhisker), height = 0.2, color = "blue") +
  geom_text(aes(label = paste0("Life Exp: ", round(`Healthy life expectancy`, 1))),
            hjust = -0.2, vjust = 0.4, color = "darkgreen", size = 3) +
  labs(
    title = "Ladder Score and Life Expectancy by Region",
    x = "Ladder Score", y = "Region"
  ) +
  theme_minimal()
```
The graph displays average Ladder Scores (a proxy for life satisfaction) across world regions, with horizontal error bars showing upper and lower confidence intervals. Regions like Western Europe, North America & ANZ, and East Asia report the highest Ladder Scores, suggesting greater average happiness levels. In contrast, Sub-Saharan Africa and South Asia rank lowest. Above each region, the corresponding Healthy Life Expectancy is labeled — regions with higher life expectancy tend to also report higher happiness scores. This visual supports the notion that health outcomes are closely associated with subjective well-being. The ordering of regions by life expectancy helps highlight this pattern clearly.



1b)

```{r fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
library(tidyverse)

# Step 1: Select relevant columns and filter South Africa
ladder_components <- combined_happy_data %>%
  select(`Country name`, `Regional indicator`, 
         starts_with("Explained by"), 
         `Dystopia + residual`) %>%
  pivot_longer(
    cols = starts_with("Explained by"):`Dystopia + residual`,
    names_to = "Component", values_to = "Value"
  )

# Step 2: Create South Africa bar as separate for first
south_africa <- ladder_components %>%
  filter(`Country name` == "South Africa")

# Step 3: Create regional average bars
regional_averages <- ladder_components %>%
  filter(`Country name` != "South Africa") %>%
  group_by(`Regional indicator`, Component) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(`Regional indicator` = fct_reorder(`Regional indicator`, Value, .fun = sum, .desc = TRUE))

# Step 4: Combine South Africa + Regions
regional_averages <- bind_rows(
  south_africa %>% 
    mutate(`Regional indicator` = "South Africa (own bar)"),
  regional_averages
)

# Step 5: Plot
ggplot(regional_averages, aes(x = `Regional indicator`, y = Value, fill = Component)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Breakdown of Ladder Score by Region",
    x = "Region", y = "Ladder Score Components"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
The plot shows the composition of the Ladder Score across global regions, broken down into components like social support, generosity, and residuals. Western Europe and North America lead with the highest scores, driven by strong contributions from GDP and social support. South Africa’s bar (added manually) appears prominently, with a high contribution from “Dystopia + residual,” indicating significant unexplained happiness. In contrast, regions like Sub-Saharan Africa and South Asia show lower scores, largely due to smaller contributions from economic and institutional factors. The comparison reveals wide disparities in the structural sources of well-being across the globe.


# Conclusion

The analysis reveals that Healthy Life Expectancy strongly correlates with higher life satisfaction scores across regions. Regions like Western Europe and North America show both high ladder scores and long life expectancies, while Sub-Saharan Africa lags in both. South Africa stands out with a relatively high Ladder Score compared to its region, largely driven by residuals and social support. The stacked barplot breakdown emphasizes that non-economic factors such as freedom and social trust play a vital role. This suggests that money alone does not guarantee happiness. Investments in health and well-being infrastructure may yield better outcomes for life satisfaction. Overall, happiness is multifaceted and context-dependent.





