---
# IMPORTANT: Change settings here, but DO NOT change the spacing.
# Remove comments and add values where applicable.
# The descriptions below should be self-explanatory

title: "A Data-Driven Exploration of Baby Naming Trends in the United States"
#subtitle: "This will appear as Right Header"

documentclass: "elsarticle"

# --------- Thesis title (Optional - set to FALSE by default).
# You can move the details below around as you please.
Thesis_FP: FALSE
# Entry1: "An unbelievable study with a title spanning multiple lines."
# Entry2: "\\textbf{Some Guy}" # textbf for bold
# Entry3: "A thesis submitted toward the degree of Doctor of Philosophy"
# Uni_Logo: Tex/Logo.png # Place a logo in the indicated location (from your root, e.g. defaults to ~/Tex/Logo.png) and uncomment this line. Leave uncommented for no image
# Logo_width: 0.3 # If using a logo - use this to set width (size) of image
# Entry4: "Under the supervision of: \\vfill Prof. Joe Smith and Dr. Frank Smith"
# Entry5: "Stellenbosch University"
# Entry6: April 2020
# Entry7:
# Entry8:

# --------- Front Page
# Comment: ----- Follow this pattern for up to 5 authors
AddTitle: TRUE # Use FALSE when submitting to peer reviewed platform. This will remove author names.
Author1: "Linda Dube"
Ref1: "Stellenbosch University, Western Cape" # First Author's Affiliation
Email1: "23084103\\@sun.ac.za" # First Author's Email address


CorrespAuthor_1: False  # If corresponding author is author 3, e.g., use CorrespAuthor_3: TRUE



# ----- Manage headers and footers:
#BottomLFooter: $Title$
#BottomCFooter:
#TopLHeader: \leftmark # Adds section name at topleft. Remove comment to add it.
BottomRFooter: "\\footnotesize Page \\thepage" # Add a '#' before this line to remove footer.
addtoprule: TRUE
addfootrule: TRUE               # Use if footers added. Add '#' to remove line.

# --------- page margins:
margin: 2.3 # Sides
bottom: 2 # bottom
top: 2.5 # Top
HardSet_layout: TRUE # Hard-set the spacing of words in your document. This will stop LaTeX squashing text to fit on pages, e.g.
# This is done by hard-setting the spacing dimensions. Set to FALSE if you want LaTeX to optimize this for your paper.

# --------- Line numbers
linenumbers: FALSE # Used when submitting to journal

# ---------- References settings:
# You can download cls format here: https://www.zotero.org/ - simply search for your institution. You can also edit and save cls formats here: https://editor.citationstyles.org/about/
# Hit download, store it in Tex/ folder, and change reference below - easy.
bibliography: Tex/ref.bib       # Do not edit: Keep this naming convention and location.
csl: Tex/harvard-stellenbosch-university.csl # referencing format used.
# By default, the bibliography only displays the cited references. If you want to change this, you can comment out one of the following:
#nocite: '@*' # Add all items in bibliography, whether cited or not
# nocite: |  # add specific references that aren't cited
#  @grinold2000
#  @Someoneelse2010

# ---------- General:
RemovePreprintSubmittedTo: TRUE  # Removes the 'preprint submitted to...' at bottom of titlepage
Journal: "Journal of Finance"   # Journal that the paper will be submitting to, if RemovePreprintSubmittedTo is set to TRUE.
toc: FALSE                       # Add a table of contents
numbersections: TRUE             # Should sections (and thus figures and tables) be numbered?
fontsize: 11pt                  # Set fontsize
linestretch: 1.2                # Set distance between lines.
link-citations: TRUE            # This creates dynamic links to the papers in reference list.

### Adding additional latex packages:
# header-includes:
#    - \usepackage{colortbl} # Add additional packages here.

output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 3.5 # Adjust default figure sizes. This can also be done in the chunks of the text.
    fig_height: 3.5
    latex_engine: xelatex
abstract: |
 This study employs Spearman rank correlation analysis to examine longitudinal trends in U.S. baby naming practices from 1910 to 2014. By comparing annual rankings of the top 20 male and female names across consecutive years, we assess the temporal persistence of naming conventions. The results indicate significant transformations in naming patterns during the 1990s, reflecting evolving cultural influences on parental decision-making. These findings contribute to our understanding of how sociocultural factors shape intergenerational naming traditions in American society.
---

<!-- First: Set your default preferences for chunk options: -->

<!-- If you want a chunk's code to be printed, set echo = TRUE. message = FALSE stops R printing ugly package loading details in your final paper too. I also suggest setting warning = FALSE and checking for warnings in R, else you might find ugly warnings in your paper. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.pos="H", fig.pos = 'H')
# Note: Include = FALSE implies the code is executed, but not printed in your pdf.
# warning and message = FALSE implies ugly messages and warnings are removed from your pdf.
# These should be picked up when you execute the command chunks (code sections below) in your rmd, not printed in your paper!

# Lets load in example data, and see how this can be stored and later called from your 'data' folder.
if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
Example_data <- Texevier::Ex_Dat

# Notice that as you are working in a .Rproj file (I am assuming you are) - the relative paths of your directories start at your specified root.
# This means that when working in a .Rproj file, you never need to use getwd() - it is assumed as your base root automatically.
write_rds(Example_data, path = "data/Example_data.rds")

```


<!-- ############################## -->
<!-- # Start Writing here: -->
<!-- ############################## -->

# Introduction \label{Introduction}

This project investigates the dynamic evolution of baby naming conventions in the United States from the early 20th century to present day, analyzing how cultural, musical, and media influences have shaped naming trends. Using longitudinal data from Social Security records, Billboard charts, and entertainment media, I employ quantitative methods to identify patterns and disruptions in naming persistence. The study particularly examines the growing volatility in name popularity since the 1990s, contrasting traditional naming practices with modern, media-influenced choices. Through this analysis, I aim to illuminate the complex interplay between cultural phenomena and parental naming decisions across generations.

# Data and method {-}

This study leverages four comprehensive datasets to analyze cultural influences on baby naming trends in the United States. The primary dataset consists of official Social Security Administration records documenting the annual popularity of baby names through 2015, providing a foundation for tracking naming patterns over time. To examine potential pop culture influences, we incorporate Billboard Hot 100 chart data spanning 1958-2015, which tracks weekly song and artist rankings based on sales, radio play, and streaming metrics. Additionally, we utilize movie and television data from HBO/TMDB, including titles, release years, and popularity scores, to identify correlations between character names and baby name trend



```{r}
library(tidyverse)
list.files("C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/23084103_Datascience_Exam/Tex_Ex/23084103_Exam/code/Question1_Plots.R", full.names = T, recursive = T) %>% as.list() %>% walk(~source(.))


baby_names <- readRDS("C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/23084103_Datascience_Exam/Tex_Ex/23084103_Exam/Question 1/Tex_Ex/23034103_Q1USbabynames/data/US_Baby_names/Baby_Names_By_US_State.rds")

billboard <- readRDS("C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/23084103_Datascience_Exam/Tex_Ex/23084103_Exam/Question 1/Tex_Ex/23034103_Q1USbabynames/data/US_Baby_names/charts.rds")

hbo_titles <- readRDS("C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/23084103_Datascience_Exam/Tex_Ex/23084103_Exam/Question 1/Tex_Ex/23034103_Q1USbabynames/data/US_Baby_names/HBO_titles.rds")

hbo_credits <- readRDS("C:/Users/sukol/Documents/Masters frst semester/Data Science Exam/Data-Science-Exam/23084103_Datascience_Exam/Tex_Ex/23084103_Exam/Question 1/Tex_Ex/23034103_Q1USbabynames/data/US_Baby_names/HBO_credits.rds")

library(skimr)

skim(baby_names)

```

The dataset contains over 5.6 million baby name records from all U.S. states between 1910 and 2014, spread across six variables—three character and three numeric. Names are consistently recorded, with no missing values and over 30,000 unique entries. The distribution of years is broad and fairly even, while name counts per entry vary widely, suggesting a skew toward rare names. The dataset is clean, complete, and well-suited for time-series and trend analysis of naming patterns across gender, time, and state.

<!-- The following is a code chunk. It must have its own unique name (after the r), or no name. After the comma follows commands for R which are self-explanatory. By default, the code and messages will not be printed in your pdf, just the output: -->
\newpage


Findings
#  Spearman Rank Correlation of Baby Names


```{r Figure1, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}

library(dplyr)
library(ggplot2)

correlations <- baby_names %>%  
  group_by(Year, Gender) %>%
  summarise(
    Spearman = cor(rank(Count), rank(lag(Count)), 
                 method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )
 # before 1990
 avg_correlation <- correlations %>%
     group_by(Year, Gender) %>%
     summarise(Avg_Spearman = mean(Spearman, na.rm = TRUE), .groups = "drop")
 
 ggplot(avg_correlation, aes(x = Year, y = Avg_Spearman, color = Gender)) +
     geom_line(size = 1.2) +
     geom_point(alpha = 0.6) +
     geom_vline(xintercept = 1910, linetype = "dashed", color = "red") +
     labs(
         title = "Name Persistence by Gender: Average Spearman Correlation of Top 25 Names",
         x = "Year",
         y = "Average Spearman Rank Correlation",
         color = "Gender"
     ) +
     theme_minimal()
  
 
 
 avg_correlation <- correlations %>%
     filter(Year <= 1990) %>%   # only include years after 1990
     group_by(Year, Gender) %>%
     summarise(
         Avg_Spearman = mean(Spearman, na.rm = TRUE),
         .groups = "drop"
     )
 
 ggplot(avg_correlation, aes(x = Year, y = Avg_Spearman, color = Gender)) +
     geom_line(size = 1.2) +
     geom_point(alpha = 0.6) +
     labs(
         title = "Name Persistence by Gender (After 1990)",
         x = "Year",
         y = "Average Spearman Rank Correlation",
         color = "Gender"
     ) +
    theme_minimal()
 
```
  
The Spearman rank correlation plot below illustrates the year-on-year persistence of baby name popularity before 1990, separated by gender. The correlation for female' names was consistently high, typically above 0.9, indicating that the most popular male names changed very little from one year to the next. In contrast, the correlation for boys' names was generally lower and more volatile, reflecting greater fluctuations in naming trends. This suggests that female names were more susceptible to changing cultural influences, while male names remained relatively traditional and stable over time. Despite the fluctuations, both genders show a moderate-to-high level of persistence overall.
  
  
  

```{r Figure 1a, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}
 # Calculate Spearman correlations first
library(dplyr)
library(ggplot2)

correlations <- baby_names %>%  
  group_by(Year, Gender) %>%
  summarise(
    Spearman = cor(rank(Count), rank(lag(Count)), 
                 method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )
 # After 1990
 avg_correlation <- correlations %>%
     group_by(Year, Gender) %>%
     summarise(Avg_Spearman = mean(Spearman, na.rm = TRUE), .groups = "drop")
 
 ggplot(avg_correlation, aes(x = Year, y = Avg_Spearman, color = Gender)) +
     geom_line(size = 1.2) +
     geom_point(alpha = 0.6) +
     geom_vline(xintercept = 1910, linetype = "dashed", color = "red") +
     labs(
         title = "Name Persistence by Gender: Average Spearman Correlation of Top 25 Names",
         x = "Year",
         y = "Average Spearman Rank Correlation",
         color = "Gender"
     ) +
     theme_minimal()
  
 
 
 avg_correlation <- correlations %>%
     filter(Year > 1990) %>%   # only include years after 1990
     group_by(Year, Gender) %>%
     summarise(
         Avg_Spearman = mean(Spearman, na.rm = TRUE),
         .groups = "drop"
     )
 
 ggplot(avg_correlation, aes(x = Year, y = Avg_Spearman, color = Gender)) +
     geom_line(size = 1.2) +
     geom_point(alpha = 0.6) +
     labs(
         title = "Name Persistence by Gender (After 1990)",
         x = "Year",
         y = "Average Spearman Rank Correlation",
         color = "Gender"
     ) +
    theme_minimal()
 
```
The Spearman rank correlation after 1990 shows slightly more fluctuation than in earlier decades, particularly for female names, which dip as low as 0.75. While male names remain relatively stable, they also show occasional drops below 0.85. This suggests that name trends have become less persistent and more influenced by short-term cultural factors. Thus, I then confirm the suspicion that name popularity has become more volatile since the 1990s.
 
 
 # Year-on-Year Surges in Baby Name Popularity


```{r Figure 1b, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}

 library(dplyr)
 # Get top 500 names to reduce noise
 top_names <- baby_names %>%
     group_by(Year, Gender) %>%
     arrange(desc(Count)) %>%
     slice_head(n = 500) %>%
     ungroup()
 
 # Calculate year-on-year % change in count
 popularity_change <- top_names %>%
     arrange(Gender, Name, Year) %>%
     group_by(Gender, Name) %>%
     mutate(
         Count_lag = lag(Count),
         Pct_Change = (Count - Count_lag) / Count_lag
     ) %>%
     ungroup() %>%
     filter(!is.na(Pct_Change), Count_lag > 20)  # remove low base counts
 
 # Top 20 surges overall
 biggest_surges <- popularity_change %>%
     arrange(desc(Pct_Change)) %>%
     slice_head(n = 20)
 
 print(biggest_surges)
``` 
```{r Figure 2a, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}

library(ggplot2)

# Example: plot top surging names
plot_names <- c("John", "Mary", "Joseph", "William")

baby_names %>%
  filter(Name %in% plot_names) %>%
  ggplot(aes(x = Year, y = Count, color = Name)) +
  geom_line(size = 1.2) +
  labs(
    title = "Surges in Name Popularity Over Time",
    x = "Year", y = "Count"
  ) +
  theme_minimal(base_size = 13)
```
The graph highlights four names—John, Mary, William, and Joseph—that showed distinct surges in popularity over time.
Mary and John peaked early, especially between the 1910s and 1950s, reflecting their strong biblical and traditional appeal.
William remained relatively stable, while Joseph saw renewed interest around the 1990s–2000s, possibly tied to cultural or religious revivals.
Their early “pop” suggests these names were historically dominant and shaped broader naming trends across decades.

\newpage

# Checking for baby name spikes after major events
```{r Figure 3a, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}
# Example: Find baby names that match artist names from Billboard
library(dplyr)

# Get unique artist names from Billboard data (assuming first word as first name)
artist_names <- billboard %>%
  mutate(artist_first_name = word(artist, 1)) %>%
  distinct(artist_first_name) %>%
  pull(artist_first_name)

# Find matching baby names
name_trends <- baby_names %>%
  filter(Name %in% artist_names) %>%
  group_by(Name, Year) %>%
  summarize(Total = sum(Count))

# Visualize
ggplot(name_trends, aes(x = Year, y = Total, color = Name)) +
  geom_line() +
  labs(title = "Baby Names Matching Music Artists Over Time")


# Example: Whitney Houston's debut (1985)
Kanye_data <- baby_names %>%
  filter(Name == "Kanye") %>%
  mutate(After1985 = Year >= 1995)

ggplot(Kanye_data, aes(x = Year, y = Count, fill = After1985)) +
  geom_col() +
  geom_vline(xintercept = 1985, linetype = "dashed") +
  labs(title = "'Kanye' Names Before/After Kanye West's Fame")
```
The graph shows a dramatic rise in babies named "Kanye" starting in 2001, coinciding exactly with Kanye West's breakout year after his debut album The College Dropout and work on Jay-Z's The Blueprint. This 1-year lag reflects the natural delay between cultural exposure (parents hearing the name in 2001) and birth registration. The sustained popularity post-2001 suggests Kanye's continued fame reinforced the name's appeal, while its rarity before 2001 confirms the artist's direct influence. Like "Whitney" in the 1980s, this mirrors how distinctive names spike when associated with rising stars, proving pop culture's power to reshape naming trends almost overnight.


# Character Name Analysis (HBO_credits Data)

```{r Figure 4a, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}
# Get first names from actor credits

actor_names <- hbo_credits %>%
  mutate(actor_first_name = word(name, 1)) %>%
  distinct(actor_first_name, id)
# First, properly extract the Game of Thrones release year
game_of_thrones_release <- hbo_titles %>%
  filter(str_detect(title, "Game of Thrones")) %>%
  pull(release_year) %>%
  first()  # Takes the first release year if multiple entries exist

# Now analyze the baby name trend
arya_trend <- baby_names %>%
  filter(Name == "Arya") %>%  # Changed from "Mary" to "Arya" for relevance
  mutate(PostGOT = Year >= game_of_thrones_release)

# Plot with vertical line at release year
ggplot(arya_trend, aes(x = Year, y = Count)) +
  geom_col(aes(fill = PostGOT)) +
  geom_vline(xintercept = game_of_thrones_release, 
             linetype = "dashed", color = "red") +
  labs(title = "'Arya' Names Before/After Game of Thrones (2011)",
       subtitle = "Vertical line marks series premiere",
       x = "Year", y = "Number of Babies") +
  scale_fill_manual(values = c("FALSE" = "gray70", "TRUE" = "#1a64d6"),
                    name = "Post-Game of Thrones") +
  theme_minimal()
```
The graph tracking the name "Arya" before and after the 2011 premiere of Game of Thrones reveals a striking example of pop culture's influence on baby naming trends. Prior to 2011, the name was virtually nonexistent, with near-zero recorded births, highlighting its obscurity before the series debuted. However, a dramatic spike occurs immediately after 2011, coinciding with the show's rise to global fame and the introduction of Arya Stark as a beloved, strong-willed character. The graph serves as compelling evidence that fictional characters—especially those with strong, positive associations—can significantly shape real-world naming practices.

# Top 10 famous artists

```{r Figure 5a, warning =  FALSE, fig.align = 'center', fig.cap = "Caption Here \\label{Figure1}", fig.height = 3, fig.width = 6, dev = 'png'}
songs_data <- data.frame(billboard)


# Group by artist and count occurrences
artist_counts <- songs_data %>%
    group_by(artist) %>%
    summarise(count = n())

# Select top 10 artists based on count
top10_artists <- artist_counts %>%
    top_n(10, count)  # Select top 10 artists by count

# Plotting
ggplot(top10_artists, aes(x = reorder(artist, count), y = count)) +
    geom_bar(stat = "identity", fill = "green") +
    labs(x = "Artist", y = "Number of Songs", title = "Top 10 Popular Artists") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
#  Bubble plot baby names counts by year
```{r}
aggregate_names <- function(baby_names) {
    baby_names %>%
        group_by(Year, Name) %>%
        summarise(Count = sum(Count)) %>%
        arrange(desc(Count))
}

# Aggregate baby names data
agg_baby_names <- aggregate_names(baby_names)

# Select the top 10 names per year and ungroup the data
top_names_per_year <- agg_baby_names %>%
    group_by(Year) %>%
    slice_max(Count, n = 1) %>%
    ungroup()

# Create the bubble plot with adjusted transparency and size scale
bubble_plot <- ggplot(top_names_per_year, aes(x = Year, y = Name, size = Count)) +
    geom_point(alpha = 0.6, color = "powderblue") +  # Adjust color and transparency
    scale_size_continuous(range = c(2, 15)) +  # Adjust size range as needed
    labs(
        title = "Popularity of Baby Names Over Time (Top 10 Names Per Year)",
        x = "Year",
        y = "Name",
        size = "Count"  # Legend title for the size aesthetic
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7)
    )

# Display the plot
print(bubble_plot)


```

This bubble plot visualizes the popularity trends of the top 10 baby names in the U.S. over time, with each bubble representing a name's annual frequency. Larger bubbles (e.g., Michael, James) indicate higher name counts, showing dominant traditional names, while smaller bubbles (e.g., Isabella, Sophia) reflect more modern favorites. The x-axis (Year) reveals shifts in naming trends, such as Mary and Linda peaking mid-century before declining, while Emma and Jacob surged in recent decades. The plot highlights how naming preferences evolve, blending enduring classics with newer influences.

# Conclusion

Traditional names like John and Mary dominated early eras with stable, biblical appeal, while modern naming trends—especially post-1990—show increasing volatility, particularly for female names. The dramatic spikes in "Kanye" (2001) and "Arya" (2011) prove that pop culture (music, TV) can rapidly transform obscure names into mainstream choices, often with a 1–2 year lag. Male names historically exhibited stronger persistence, but recent decades reveal a cultural shift where all names are now more susceptible to short-term influences. Together, these patterns highlight how naming trends evolved from tradition-bound stability to dynamic reflections of media, celebrity, and societal change.




